---
title: "Reproducible Research: Peer Assessment 1"
author: "Hirley Dayan"
date: "May 16, 2015"
output: html_document
---

###Loading and preprocessing the data

For performing the analysis presented in this report the data can be download 
from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Initially all the needed libraries are loaded using the lines:

```{r}
library(data.table)
library(plyr)
library(ggplot2)
```

The following lines autommatically download the file into the working directory, 
unzip it and delete the zip file for keeping the workding directory clean.

```{r, echo = TRUE}
file.src = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
file.zip = "repdata_data_activity.zip"
dwmethod = "curl"
download.file(url = file.src, destfile = file.zip, method = dwmethod)
unzip(file.zip)
file.remove(file.zip)
```

The unzipped file, which is in csv file, is read with the commands in the next 
lines.

```{r, echo=TRUE}
file.csv = "activity.csv"
activity.dt <- data.table(read.csv(file.csv))
```

The variable data is formated accordingly using the following code line:

```{r}
activity.dt$date <- as.Date(activity.dt$date, format = "%Y-%m-%d")
```

A short printout of the loaded dataset is shown following:

```{r}
head(activity.dt)
```

The commands following cleans the dataset removing any NA entrie, creating 
a new dataset without missing values.

```{r, echo=TRUE}
activity.dt.clean <- activity.dt[!is.na(activity.dt$steps)]
head(activity.dt.clean)
na.number <- nrow(activity.dt.clean[is.na(activity.dt.clean$steps)])
```

The number of **missing values** in the clean database is **`r na.number`**.

###What is mean total number of steps taken per day?

The daily total amount of steps is calculated in the next lines.

```{r, echo=TRUE}
steps.daily <- aggregate(activity.dt.clean$steps, 
                               by=list(date = activity.dt.clean$date), 
                               FUN=sum)
setnames(steps.daily,"x","steps")
head(steps.daily)
````

The daily total amount of steps is shown by the graph following:

```{r, echo=TRUE}
qplot(steps.daily[!is.na(steps.daily$steps),]$steps, 
      xlab = "Total daily steps distribution",
      ylab = "Frequency",
      main = "Histogram of Total Daily Steps") +
        geom_histogram(aes(fill = ..count..)) +
        scale_fill_gradient("Count", low = "green", high = "red")
```

The daily mean and median amounts of steps are shown in the next lines.

```{r, echo=TRUE}
steps.mean <- mean(steps.daily$steps)
steps.median <- median(steps.daily$steps)
```

The mean is **`r format(steps.mean,scientific = FALSE)`** and the median is 
**`r format(steps.median, scientific = FALSE)`**.

###What is the average daily activity pattern?

The average daily activity pattern is shown in the graph following:

```{r, echo=TRUE}
steps.interval.mean <- aggregate(activity.dt.clean$steps, 
                                by = list(interval = activity.dt.clean$interval),
                                FUN=mean)
setnames(steps.interval.mean,"x", "steps")
ggplot(data=steps.interval.mean, aes(x=interval, y=steps)) + 
        geom_line(colour="red", size=1) + 
        labs(title="Average Daily Activity", x="Interval", y="Steps")
```

###Imputing missing values

```{r, echo=TRUE}
na.number <- nrow(activity.dt[is.na(activity.dt$steps)])
```

A number of **`r na.number`** entries in the dataset has **missing values**.

For filling the gaps in the dataset, the commands shown in the lines following 
replace the NA steps with the mean steps values, in integer, from the equivalent 
interval .

```{r, echo=TRUE}
na_replacement.dt <- merge(activity.dt[is.na(activity.dt$steps)],
                           steps.interval.mean,by="interval",
                           , all = TRUE)
na_replacement.dt <- na_replacement.dt[,.(interval = interval,
                                       date = date,
                                       steps = steps.y)]

activity.complete.dt <- merge(activity.dt,
                              na_replacement.dt,
                              by = c("date", "interval"), 
                              all = TRUE)

activity.complete.dt[is.na(activity.complete.dt$steps.x)] <- 
        mutate(activity.complete.dt[is.na(activity.complete.dt$steps.x)],
               steps.x = as.integer(steps.y))

activity.complete.dt <- activity.complete.dt[,.(steps = steps.x,
                                                date = date, 
                                                interval = interval)]
na.number <- nrow(activity.complete.dt[is.na(activity.complete.dt$steps)])
```

The number of **missing values** in the database after the steps above is 
**`r na.number`**.

```{r, echo=TRUE}
head(activity.complete.dt)
```

The daily total amount of steps, using the adjusted ***missing values*** dataset,
is shown by the graph following:

```{r, echo=TRUE}
steps.daily <- aggregate(activity.complete.dt$steps, 
                               by=list(date = activity.complete.dt$date), 
                               FUN=sum)
setnames(steps.daily,"x","steps")
head(steps.daily)

qplot(steps.daily[!is.na(steps.daily$steps),]$steps, 
      xlab = "Total daily steps distribution",
      ylab = "Frequency",
      main = "Histogram of Total Daily Steps") +
        geom_histogram(aes(fill = ..count..)) +
        scale_fill_gradient("Count", low = "green", high = "red")
```

The daily mean and median amounts of steps are shown in the next lines.

```{r, echo=TRUE}
steps.mean <- mean(steps.daily$steps)
steps.median <- median(steps.daily$steps)
```

The mean is **`r format(steps.mean,scientific = FALSE)`** and the median is 
**`r format(steps.median, scientific = FALSE)`**.

###Are there differences in activity patterns between weekdays and weekends?

The code lines following calculate the activity patterns during the weekdays 
and weekends.

```{r, echo=TRUE}
activity.week.dt <- mutate(activity.complete.dt,
                               weekday = weekdays(activity.complete.dt$date))

weekends <- c("Saturday","Sunday")
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity.week.dt[activity.week.dt$weekday %in% weekends]$weekday <- "Weekend"
activity.week.dt[activity.week.dt$weekday %in% weekdays]$weekday <- "Weekday"

activity.week.dt$weekday <- as.factor(activity.week.dt$weekday)

activity.weekend.dt <- activity.week.dt[activity.week.dt$weekday == "Weekend"]
steps.weekend <- aggregate(activity.weekend.dt$steps, 
                               by=list(interval = activity.weekend.dt$interval), 
                               FUN=mean)
steps.weekend$weekday <- "Weekend"
setnames(steps.weekend,"x", "steps")

activity.weekday.dt <- activity.week.dt[activity.week.dt$weekday == "Weekday"]
steps.weekday <- aggregate(activity.weekday.dt$steps, 
                               by=list(interval = activity.weekday.dt$interval), 
                               FUN=mean)
steps.weekday$weekday <- "Weekday"
setnames(steps.weekday,"x", "steps")

steps.week <- rbind(steps.weekday,steps.weekend)
```

The graph following shows the mean number of steps during the analyzed dataset 
period during the weekdays and weekends.

```{r, echo=TRUE}
ggplot(steps.week, aes(x=interval, y=steps)) + 
        geom_line(color="blue") + 
        facet_wrap(~ weekday, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```
